---
title: "Basic Digital Out Functionality"
author: "Andrey Samokhin"
date: "2026-09-22"
output:
  rmarkdown::html_vignette:
    css: style.css
vignette: >
  %\VignetteIndexEntry{Basic Digital Out Functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.height = 3.5,
  fig.width = 5.5
)
```

```{r session_setup, include = FALSE}
library(dwf4r)
package_info <- paste0("`dwf4r_", utils::packageVersion("dwf4r"), "`")
```

## Experimental setup

The examples in this vignette were recorded using an Analog Discovery 2 NI
Edition and a Rigol DHO804 digital oscilloscope.

The examples and descriptions were checked against WaveForms SDK 3.25 and the
WaveForms SDK Reference Manual revised March 4, 2026.

### Pattern generator

The examples were tested using a Digilent Analog Discovery 2 NI Edition.

The device is opened with `OpenDevice()`, used throughout the vignette, and
closed after the final example. Its information is shown below; the serial
number has been masked because it is not relevant to reproducing the examples.

When exactly one device is available, `OpenDevice()` selects it automatically;
if several devices are connected, specify the required device by serial number.

```{r open_device, eval = FALSE}
device <- OpenDevice()

print(device$info)

#> $device_type
#> [1] "Analog Discovery 2"
#>
#> $device_revision
#> [1] 2051
#>
#> $device_name
#> [1] "Analog Discovery 2 NI EditionSN:Discovery2NI"
#>
#> $user_name
#> [1] "Discovery2NI"
#>
#> $serial_number
#> [1] "SN:************"
#>
#> $config_count
#> [1] 8
```

### Oscilloscope

The waveforms shown in this vignette were measured using a Rigol DHO804
oscilloscope. Up to four Digital Out channels were connected to four
oscilloscope inputs, and the Analog Discovery ground terminal was connected to
the corresponding oscilloscope grounds.

For each example, the waveform was generated using the displayed R code, the
oscilloscope settings were adjusted to show the relevant signal properties, and
a screenshot was saved for inclusion in the vignette. The R code is not
evaluated during vignette rendering because it requires connected hardware.

## Generate a pulse waveform

Among the available Digital Out signal-generation functions, `"pulse"` is one of
the simplest and most generally useful. It produces a periodic digital signal
that alternates between low and high output states.

The signal-generation function is selected with `SetDigitalOutFunction()`. The
WaveForms SDK refers to this setting as the Digital Out *type*, whereas `dwf4r`
uses the term *function* to maintain consistent naming between Analog Out and
Digital Out.

Digital Out also supports several electrical output modes. In `"push_pull"`
mode, the output actively drives both the low and high states. In `"open_drain"`
mode, the output actively drives the low state and releases the output for the
high state, whereas `"open_source"` mode actively drives the high state and
releases the output for the low state. The `"three_state"` mode additionally
permits a high-impedance output state. The modes supported by a particular
channel depend on the device and can be queried with
`GetDigitalOutOutputModes()`. Open-drain and open-source operation requires an
appropriate external pull resistor; three-state output is available only with
supported signal-generation functions. The examples in this vignette use
`"push_pull"`, which directly produces the low and high voltage levels required
for the demonstrated pulse waveforms.

Unlike Analog Out, Digital Out pulse frequency and duty cycle are not specified
directly. Instead, they are determined by three whole-number values: a clock
divider and low and high counters. The divider reduces the Digital Out internal
clock frequency and determines the duration of one counter step:

$$
t_{\mathrm{count}} = \frac{ \mathrm{divider} }{ f_{\mathrm{clock}} }.
$$

The low and high counters specify how many counter steps the output remains in
the corresponding state. Consequently, the pulse frequency is

$$
f_{\mathrm{pulse}} =
\frac{ f_{\mathrm{clock}} }
{ \mathrm{divider} \cdot (N_{\mathrm{low}} + N_{\mathrm{high}}) },
$$

and the fraction of each period spent in the high state is

$$
D = \frac{ N_{\mathrm{high}} }{ N_{\mathrm{low}} + N_{\mathrm{high}} }.
$$

The following example uses these settings to generate a continuous 1 kHz pulse
waveform with a 50% duty cycle on the first Digital Out channel.

```{r generate_pulse, eval = FALSE}
ResetDigitalOut(device)

EnableDigitalOutChannel(device, channel = 0L)
SetDigitalOutFunction(device, channel = 0L, func = "pulse")
SetDigitalOutOutput(device, channel = 0L, output = "push_pull")

frequency <- 1000
n_counts <- 10L
internal_clock <- GetDigitalOutInternalClockFrequency(device)
divider <- round(internal_clock / (n_counts * frequency))
SetDigitalOutDivider(device, channel = 0L, divider = divider)

SetDigitalOutCounter(
  device,
  channel = 0L,
  low_count = 5L,
  high_count = 5L
)

StartDigitalOut(device)
```

The example begins with `ResetDigitalOut()` to restore the Digital Out
instrument to its SDK defaults and provide a reproducible starting point. The
first channel is then enabled and configured for push-pull pulse generation.
After the timing parameters are set, `StartDigitalOut()` starts the configured
Digital Out instrument.

In this example, one pulse period consists of ten counter steps. The calculated
divider makes these ten steps correspond to a frequency of 1 kHz. Five steps are
assigned to each output state, producing a 50% duty cycle. According to the
Analog Discovery 2 specification, its Digital I/O uses 3.3 V CMOS logic. The
measured push-pull waveform shows the corresponding low and high logic levels
(Figure 1).

```{r}
#| label: fig_pulse_waveform
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 1. Oscilloscope trace of the generated 1 kHz pulse waveform with a
#|   50% duty cycle on Digital Out channel 0.

knitr::include_graphics(
  "figures/digital-out-basics/pulse-waveform.png"
)
```

## Modify a running waveform

Digital Out parameters can also be changed while the instrument is running.
Because automatic configuration is disabled by `dwf4r`, setting a parameter
updates the pending configuration without immediately changing the generated
signal. The updated configuration is applied when `StartDigitalOut()` is called,
which starts or restarts the Digital Out instrument.

The following code changes the pulse duty cycle from 50% to 20% while retaining
the 1 kHz frequency.

```{r modify_running_waveform, eval = FALSE}
SetDigitalOutCounter(
  device,
  channel = 0L,
  low_count = 8L,
  high_count = 2L
)
StartDigitalOut(device)
```

Unlike Analog Out, Digital Out does not provide a separate operation equivalent
to `ApplyAnalogOutSettings()`. Calling `StartDigitalOut()` therefore applies the
pending configuration by restarting the instrument.

Only the low and high counters are changed; the clock divider and other channel
settings retain their previous values. The total number of counter steps remains
ten, so the pulse frequency is unchanged. Eight steps are assigned to the low
state and two to the high state, producing a 20% duty cycle (Figure 2).

```{r}
#| label: fig_pulse_waveform_20_percent
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 2. Oscilloscope trace after changing the 1 kHz pulse waveform from
#|   50% to 20% duty cycle. The pulse period remains unchanged.

knitr::include_graphics(
  "figures/digital-out-basics/pulse-waveform-20-percent.png"
)
```

## Inspect and stop waveform generation

The current state of the Digital Out instrument can be retrieved using
`GetDigitalOutStatus()`. At this point, the instrument is generating the
modified pulse waveform and is therefore in the `"running"` state.

```{r get_digital_out_status, eval = FALSE}
GetDigitalOutStatus(device)

#> [1] "running"
```

Signal generation can be stopped using `StopDigitalOut()`. Stopping the
instrument does not reset its configuration; the configured signals can be
started again using `StartDigitalOut()`.

```{r stop_digital_out, eval = FALSE}
StopDigitalOut(device)
GetDigitalOutStatus(device)

#> [1] "ready"
```

The `"ready"` state confirms that the Digital Out instrument is no longer
generating signals.

## Generate a finite waveform burst

Digital Out generation proceeds through several instrument states (Figure 3).
The diagram is a simplified adaptation of the state model presented in the
Digilent WaveForms SDK Reference Manual. Digital Out differs from Analog Out in
that a single state machine controls all Digital Out channels.

```{r}
#| label: fig_digital_out_states
#| echo: false
#| fig.cap: >
#|   Figure 3. Simplified Digital Out state diagram. Solid arrows show the
#|   primary state sequence; dashed arrows show the repeat paths to
#|   <code>&quot;armed&quot;</code> when repeat triggering is enabled and to
#|   <code>&quot;wait&quot;</code> when it is disabled.

local({
  old_par <- par(c("mar", "xpd"))
  on.exit(par(old_par), add = TRUE)
  par(
    mar = c(0.5, 0.5, 0.5, 0.5),
    xpd = NA
  )

  state_x <- c(
    "ready" = 1,
    "armed" = 3,
    "wait" = 5,
    "done" = 1,
    "repeat" = 3,
    "running" = 5
  )
  state_y <- c(
    "ready" = 1,
    "armed" = 1,
    "wait" = 1,
    "done" = 0,
    "repeat" = 0,
    "running" = 0
  )

  box_width <- 1
  box_height <- 0.5

  plot.new()
  plot.window(
    xlim = c(0, 6),
    ylim = c(-0.5, 1.5),
    xaxs = "i",
    yaxs = "i"
  )

  rect(
    xleft = state_x - box_width / 2,
    ybottom = state_y - box_height / 2,
    xright = state_x + box_width / 2,
    ytop = state_y + box_height / 2,
    col = "gray95",
    border = "gray30",
    lwd = 1.5
  )
  
  text(
    x = state_x,
    y = state_y,
    labels = names(state_x)
  )

  arrows(
    x0 = c(
      state_x[["ready"]] + box_width / 2,
      state_x[["armed"]] + box_width / 2,
      state_x[["wait"]],
      state_x[["running"]] - box_width / 2,
      state_x[["repeat"]] - box_width / 2,
      state_x[["repeat"]],
      state_x[["repeat"]] + box_width / 2
    ),
    y0 = c(
      state_y[["ready"]],
      state_y[["armed"]],
      state_y[["wait"]] - box_height / 2,
      state_y[["running"]],
      state_y[["repeat"]],
      state_y[["repeat"]] + box_height / 2,
      state_y[["repeat"]] + box_height / 2
    ),
    x1 = c(
      state_x[["armed"]] - box_width / 2,
      state_x[["wait"]] - box_width / 2,
      state_x[["running"]],
      state_x[["repeat"]] + box_width / 2,
      state_x[["done"]] + box_width / 2,
      state_x[["armed"]],
      state_x[["wait"]] - box_width / 2
      ),
    y1 = c(
      state_y[["armed"]],
      state_y[["wait"]],
      state_y[["running"]] + box_height / 2,
      state_y[["repeat"]],
      state_y[["done"]],
      state_y[["armed"]] - box_height / 2,
      state_y[["wait"]] - box_height / 2
    ),
    length = 0.08,
    lty = c(rep("solid", 5L), rep("dashed", 2L))
  )
  
  text(
    x = c(
      (state_x[["ready"]] + state_x[["armed"]]) / 2,
      (state_x[["armed"]] + state_x[["wait"]]) / 2
    ),
    y = c(state_y[["ready"]], state_y[["armed"]]) + 0.1,
    labels = c("start", "trigger")
  )
})
```

Before generation starts, the instrument is in the `"ready"` state. Starting
Digital Out moves it through `"armed"` and then to `"wait"` under the default
configuration used here. The instrument remains in `"wait"` for the configured
wait time and then generates all enabled channels during the configured run
time. If repetitions remain, the wait-run cycle begins again; otherwise, the
instrument enters `"done"`.

By default, the wait time is `0`, the run time is `0`, and the repeat count is
`1L`. A run time of `0` requests continuous generation, so the repeat count is
not reached. With a finite run time, the repeat count specifies how many
wait-run cycles are performed; `0L` requests infinite repetition.

The pulse waveform is generated continuously while the instrument is in the
`"running"` state. A finite pulse burst is therefore produced by limiting the
Digital Out run time. The following example sets the run time to generate three
pulses.

```{r generate_three_pulse_burst, eval = FALSE}
SetDigitalOutRun(device, run_time = 3L / frequency)
StartDigitalOut(device)
```

At 1 kHz, one pulse period is 1 ms, so the configured run time is 3 ms (Figure
4). The repeat count remains at its default value of `1L`, and the run interval
is therefore performed once.

```{r}
#| label: fig_three_pulse_burst
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 4. Oscilloscope trace of a three-pulse, 1 kHz burst generated during
#|   a single 3 ms run interval.

knitr::include_graphics(
  "figures/digital-out-basics/three-pulses.png"
)
```

The progress of a sufficiently long run interval can be monitored using
`GetDigitalOutRemainingRun()`. The following example starts a one-second run,
waits for 0.1 seconds, and retrieves the remaining run value.

```{r inspect_remaining_run, eval = FALSE}
SetDigitalOutRun(device, run_time = 1)
StartDigitalOut(device)
Sys.sleep(0.1)

remaining_run_value <- GetDigitalOutRemainingRun(device)
remaining_run_value / GetDigitalOutInternalClockFrequency(device)

#> [1] 0.8450741
```

The WaveForms SDK documentation does not specify the units returned by
`FDwfDigitalOutRunStatus()`. On the Analog Discovery 2 used for this vignette,
the returned value was consistently found to be expressed in Digital Out
internal-clock ticks. Dividing it by the internal clock frequency therefore
gives the remaining run time in seconds. This behavior has not been verified for
other devices.

The converted value is expected to be slightly below the theoretical value of
0.9 seconds because the elapsed time also includes R execution, device
communication, and the status query performed by `GetDigitalOutRemainingRun()`.
This function is therefore most useful for monitoring relatively long run
intervals rather than precise short-duration timing.

Similarly, `GetDigitalOutRemainingRepeat()` retrieves the number of repeat
cycles that have not yet been completed.

```{r inspect_remaining_repeat, eval = FALSE}
SetDigitalOutTiming(
  device,
  run_time = 1 / frequency,
  repeat_count = 1000L
)
StartDigitalOut(device)
Sys.sleep(0.1)

GetDigitalOutRemainingRepeat(device)

#> [1] 842
```

In the next example, each run interval lasts for one pulse period and is
preceded by a wait interval lasting two periods. The wait-run cycle is performed
four times.

```{r generate_four_repeated_pulses, eval = FALSE}
SetDigitalOutTiming(
  device,
  run_time = 1 / frequency,
  wait_time = 2 / frequency,
  repeat_count = 4L
)
StartDigitalOut(device)
```

With the default configuration restored by `ResetDigitalOut()`, repeated
wait-run cycles proceed automatically. The output therefore consists of four
single-pulse bursts, each preceded by a 2 ms wait interval (Figure 5).

```{r}
#| label: fig_four_repeated_pulses
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 5. Oscilloscope trace of four repeated pulse bursts. Each 1 ms run
#|   interval contains one pulse period and is preceded by a 2 ms wait interval.

knitr::include_graphics(
  "figures/digital-out-basics/four-pulses.png"
)
```

## Generate a custom waveform

A custom Digital Out waveform is specified as a vector containing logical low
and high samples. Numeric patterns use `0` and `1`; character patterns can use
`"l"` and `"h"`. This demonstration uses an eight-sample pattern (Figure 6).

```{r define_custom_pattern}
pattern <- c(0, 1, 0, 0, 1, 1, 1, 1)
```

```{r}
#| label: fig_custom_pattern_profile
#| echo: false
#| fig.cap: >
#|   Figure 6. Logical profile of the eight-sample custom Digital Out pattern.
#|   Each element represents one interval at the configured sample rate.

local({
  old_par <- par(c("mar", "cex"))
  on.exit(par(old_par), add = TRUE)

  par(
    mar = c(4.1, 4.1, 1.1, 1.1),
    cex = 0.9
  )

  plot(
    seq(from = 0, to = length(pattern), by = 1),
    c(pattern, pattern[[length(pattern)]]),
    type = "s",
    xlab = "Sample interval",
    ylab = "Logical level",
    xaxs = "i",
    yaxs = "i",
    ylim = c(-0.1, 1.1),
    lwd = 2,
    yaxt = "n"
  )
  axis(2, at = c(0, 1), labels = c("low", "high"), las = 1)
})
```

To generate a custom waveform, first select the `"custom"` signal-generation
function and upload the desired pattern. Then set the `sample_rate`, which
specifies how quickly individual pattern elements are generated. It can be
viewed as the frequency applied to the individual points of the custom pattern:
at a sample rate of 8 kHz, each pattern element lasts 0.125 ms. The sample rate
therefore determines the timing of individual pattern elements, not the
repetition frequency of the complete pattern.

```{r generate_custom_waveform, eval = FALSE}
sample_rate <- 8L * 1000

SetDigitalOutFunction(device, channel = 0L, func = "custom")
SetDigitalOutCustomPattern(
  device,
  channel = 0L,
  pattern = pattern
)
SetDigitalOutCustomSampleRate(
  device,
  channel = 0L,
  sample_rate = sample_rate
)

StartDigitalOut(device)
```

The preceding timing configuration remains active: each 1 ms custom pattern is
preceded by a 2 ms wait interval, and the sequence is repeated four times
(Figure 7).

```{r}
#| label: fig_custom_pattern
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 7. Oscilloscope trace of four repetitions of the eight-sample
#|   custom Digital Out pattern. At a sample rate of 8 kHz, each pattern lasts
#|   1 ms and is preceded by a 2 ms wait interval.

knitr::include_graphics(
  "figures/digital-out-basics/custom-pattern.png"
)
```

## Control idle output

The idle mode controls a channel output while the Digital Out instrument is not
in the `"running"` state, including the `"ready"`, `"wait"`, and `"done"`
states. The available idle modes depend on the selected device and channel.

The `"initial"` mode uses the configured initial output value. The `"low"` and
`"high"` modes instead force a corresponding logical level while the instrument
is idle. The `"three_state"` mode places the output in a high-impedance state
and is not demonstrated here.

The following code sets the first channel to the `"high"` idle mode and starts
the same four-cycle custom-pattern sequence again (Figure 8).

```{r use_high_idle_output, eval = FALSE}
SetDigitalOutIdle(device, channel = 0L, idle = "high")
StartDigitalOut(device)
```

```{r}
#| label: fig_high_idle_output
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 8. Custom Digital Out pattern generated with the
#|   <code>&quot;high&quot;</code> idle mode. The output remains at the logical
#|   high level during the wait intervals and after generation finishes.

knitr::include_graphics(
  "figures/digital-out-basics/custom-pattern-idle-high.png"
)
```

## Inspect capabilities and configured values

Digital Out capabilities depend on the connected device and its selected
configuration. Before setting a parameter, `dwf4r` obtains the corresponding
capability or supported range from the WaveForms SDK and checks the requested
value against it.

The same information can also be queried explicitly. For example, the supported
divider range for a selected channel can be retrieved with:

```{r inspect_divider_range, eval = FALSE}
GetDigitalOutDividerRange(device, channel = 0L)

#> [1] 1 2147483649
```

Currently configured values can be retrieved in a similar way. For example, the
divider configured for the same channel can be inspected with:

```{r inspect_divider, eval = FALSE}
GetDigitalOutDivider(device, channel = 0L)

#> [1] 12500
```

The following functions retrieve general Digital Out capabilities:

* `GetDigitalOutChannelCount()`
* `GetDigitalOutInternalClockFrequency()`
* `GetDigitalOutRepeatRange()`
* `GetDigitalOutRunRange()`
* `GetDigitalOutWaitRange()`

The following functions retrieve capabilities of a selected Digital Out channel:

* `GetDigitalOutCounterRange()`
* `GetDigitalOutDividerRange()`
* `GetDigitalOutFunctionTypes()`
* `GetDigitalOutIdleModes()`
* `GetDigitalOutMaxDataBits()`
* `GetDigitalOutOutputModes()`

The current channel configuration can be inspected using:

* `IsDigitalOutChannelEnabled()`
* `GetDigitalOutCounter()`
* `GetDigitalOutCustomSampleRate()`
* `GetDigitalOutDivider()`
* `GetDigitalOutFunction()`
* `GetDigitalOutIdle()`
* `GetDigitalOutInitialCounter()`
* `GetDigitalOutInitialDivider()`
* `GetDigitalOutOutput()`

The instrument-level Digital Out configuration can be inspected using:

* `GetDigitalOutRepeat()`
* `GetDigitalOutRun()`
* `GetDigitalOutTiming()` (it retrieves the run time, wait time, and repeat
  count together as a named list)
* `GetDigitalOutWait()`

The current Digital Out state and remaining timing values can be inspected
using:

* `GetDigitalOutStatus()`
* `GetDigitalOutRemainingRepeat()`
* `GetDigitalOutRemainingRun()`

## Run multiple channels simultaneously

Although Digital Out generation is started and stopped for the instrument as a
whole, the waveform can be configured independently for each channel. In
particular, the signal-generation function, divider, and counter values are
channel-specific, and custom pattern data can also be defined separately for
each channel. This makes it possible to generate substantially different signals
simultaneously.

The following example configures four channels: a 0.5 kHz pulse waveform on
channel 0, two 1 kHz pulse waveforms with different duty cycles on channels 1
and 2, and a custom pattern repeating at 2 kHz on channel 3. The channels use
three different divider values, demonstrating that the divider is a
channel-specific rather than an instrument-level setting.

```{r configure_multiple_channels, eval = FALSE}
ResetDigitalOut(device)

internal_clock <- GetDigitalOutInternalClockFrequency(device)

# Use 16 counter steps per period to support 1/8-period phase shifts while
# keeping the initial divider distinct from the ordinary divider.
pulse_counts <- 16L

for (channel in seq.int(0L, 3L)) {
  EnableDigitalOutChannel(device, channel = channel)
  SetDigitalOutOutput(device, channel = channel, output = "push_pull")
}

# Channel 0: 0.5 kHz pulse waveform.
frequency_500hz <- 500
divider_500hz <- internal_clock / (pulse_counts * frequency_500hz)

SetDigitalOutFunction(device, channel = 0L, func = "pulse")
SetDigitalOutDivider(device, channel = 0L, divider = divider_500hz)
SetDigitalOutCounter(
  device,
  channel = 0L,
  low_count = (3 / 4) * pulse_counts,
  high_count = (1 / 4) * pulse_counts
)

# Channels 1 and 2: 1 kHz pulse waveforms.
frequency_1khz <- 1000
divider_1khz <- internal_clock / (pulse_counts * frequency_1khz)

SetDigitalOutFunction(device, channel = 1L, func = "pulse")
SetDigitalOutDivider(device, channel = 1L, divider = divider_1khz)
SetDigitalOutCounter(
  device,
  channel = 1L,
  low_count = (3 / 4) * pulse_counts,
  high_count = (1 / 4) * pulse_counts
)

SetDigitalOutFunction(device, channel = 2L, func = "pulse")
SetDigitalOutDivider(device, channel = 2L, divider = divider_1khz)
SetDigitalOutCounter(
  device,
  channel = 2L,
  low_count = (1 / 4) * pulse_counts,
  high_count = (3 / 4) * pulse_counts
)

# Channel 3: ten-element custom pattern repeating at 2 kHz.
pattern <- c(0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 0L)
pattern_frequency <- 2000
sample_rate <- length(pattern) * pattern_frequency
divider_custom <- round(internal_clock / sample_rate)

SetDigitalOutFunction(device, channel = 3L, func = "custom")
SetDigitalOutCustomPattern(device, channel = 3L, pattern = pattern)
SetDigitalOutDivider(device, channel = 3L, divider = divider_custom)

StartDigitalOut(device)
```

The resulting oscilloscope traces are shown in Figure 9. Because all four
signals are displayed on the same time scale, the twofold differences between
the 0.5, 1, and 2 kHz repetition frequencies can be compared directly.

```{r}
#| label: fig_multiple_channels
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 9. Oscilloscope traces of four simultaneously generated Digital Out
#|   waveforms: 0.5 kHz and 1 kHz pulse waveforms on channels 0--2 and a custom
#|   pattern repeating at 2 kHz on channel 3.

knitr::include_graphics(
  "figures/digital-out-basics/multiple-channels.png"
)
```

## Fine-tune initial waveform timing

Digital Out provides separate initial divider and counter settings for
fine-tuning the beginning of waveform generation.

By default, both the initial count and initial divider are zero. The WaveForms
SDK manual does not explicitly define the meaning of these zero values; however,
tests with the Analog Discovery 2 indicate that the corresponding initial timing
values are not applied and the ordinary counter and divider settings take effect
immediately.

The `initial_state` argument of `SetDigitalOutInitialCounter()` determines
whether a channel starts in the low or high state. With `initial_count = 0`, the
ordinary counter corresponding to the selected state is used immediately; this
behavior is consistent with Digilent's phase examples and was observed with the
Analog Discovery 2 used here. If `initial_count` is greater than zero, it
replaces the corresponding ordinary counter value for the first output-state
interval. After the initial counter expires, the ordinary low/high counter
sequence begins.

Independently of the initial counter, the initial divider configured with
`SetDigitalOutInitialDivider()` is applied to the first counter step after the
instrument enters the Running state. After this first divider interval expires,
the ordinary divider configured with `SetDigitalOutDivider()` is used for all
subsequent counter steps.

Together, these settings provide additional control over the initial timing of
individual Digital Out channels without changing their ordinary repeating
waveforms. One particularly useful application is controlling the relative phase
of signals generated on different channels. The following example uses the
previously configured pulse waveforms to demonstrate this functionality. The 1
kHz waveforms on channels 1 and 2 are given opposite phase shifts of one eighth
of their period so that their pulse centers coincide. The 0.5 kHz waveform on
channel 0 is left unchanged, so every second pair of 1 kHz pulses is also
centered with the pulse on channel 0.

There are two convenient ways to adjust the relative phase of Digital Out
waveforms. The first is to change the initial counter while leaving the initial
divider as its default value. In this case, the phase shift is
specified as a whole number of ordinary counter steps. This approach is used in
the Digital Out phase examples provided by Digilent for the C++ API. In the
following example, it is used to shift the waveform on channel 1.

An alternative is to combine the initial counter and initial divider settings.
The initial counter can be increased by one count, while the initial divider
sets the duration of the first counter step. The remaining initial counts use
the ordinary divider, so the additional duration of the first output-state
interval is determined by the initial divider. This approach can provide finer
timing than an ordinary counter step. However, an exact phase shift can be
obtained only when the required initial divider is a supported whole-number
value; otherwise, the divider must be rounded and the resulting shift is only
approximate. This approach is used below for channel 2.

```{r adjust_initial_timing, eval = FALSE}

# Channel 1: delay by 7/8 period, equivalent to a 1/8-period phase advance.
counters_ch1 <- GetDigitalOutCounter(device, channel = 1L)
SetDigitalOutInitialCounter(
  device,
  channel = 1L,
  initial_state = "low",
  initial_count = counters_ch1[["low_count"]] + (7 / 8) * pulse_counts
)

# Channel 2: delay by 1/8 period.
counters_ch2 <- GetDigitalOutCounter(device, channel = 2L)
period_ticks <- internal_clock / frequency_1khz
SetDigitalOutInitialDivider(
  device,
  channel = 2L,
  initial_divider = (1 / 8) * period_ticks
)
SetDigitalOutInitialCounter(
  device,
  channel = 2L,
  initial_state = "low",
  initial_count = counters_ch2[["low_count"]] + 1L
)

StartDigitalOut(device)
```

The resulting oscilloscope traces are shown in Figure 10. After the initial
timing adjustment, the centers of the two 1 kHz pulses coincide, and every
second pair is also centered with the 0.5 kHz pulse on channel 0.

```{r}
#| label: fig_shifted_waveforms
#| echo: false
#| out.width: "100%"
#| fig.cap: >
#|   Figure 10. Oscilloscope traces after adjusting the initial timing of
#|   channels 1 and 2. The two 1 kHz waveforms have opposite one-eighth-period
#|   phase shifts so that their pulse centers coincide; every second pair is
#|   also centered with the 0.5 kHz waveform on channel 0.

knitr::include_graphics(
  "figures/digital-out-basics/shifted-waveforms.png"
)
```

## Stop generation and close the device

At the end of the session, Digital Out generation is stopped and the device
handle is closed. `StopDigitalOut()` stops all Digital Out channels controlled
by the shared instrument state machine. `CloseDevice()` then releases the device
and discards its current configuration.

```{r close_device, eval = FALSE}
StopDigitalOut(device)
CloseDevice(device)
```


---

Built with `r package_info`.
